---
title: "Predictions"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Packages needed
library(dplyr)
library(tidyr)
library(loo)
library(cmdstanr)
library(footBayes)
library(posterior)
library(bayesplot)
library(instantiate)
library(stringi)
```

# The theory

# The data

## Data loading and preprocessing

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("ranking_preprocessing.R")
```

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("ranking2025_11_19_preprocessing.R")
```

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("results_preprocessing.R")
```

```{r}
resultat <- rbind(ranking, ranking2025_11_19)
```


## All the teams are in the ranking ?

```{r}
all_rank_norms <- unique(ranking$team_norm)
res_teams <- unique(c(results$home_norm, results$away_norm))

unmatched_in_results <- setdiff(res_teams, all_rank_norms)

cat("\n=== Teams dans results sans correspondance dans ranking ===\n")
print(sort(unmatched_in_results))
```

## All the teams in ranking are in results 

```{r}
all_rank_norms <- unique(ranking$team_norm)
res_teams <- unique(c(results$home_norm, results$away_norm))

unmatched_in_results <- setdiff(res_teams, all_rank_norms)

cat("\n=== Teams dans results sans correspondance dans ranking ===\n")
print(sort(unmatched_in_results))
```

## Correction of the names which cause problems 

```{r}
ranking <- ranking |>
  mutate(team_norm = case_when(
    
    # ── Renommages officiels ──────────────────────────────────────────────────
    team_norm == "CZECHIA"              ~ "CZECHREPUBLIC",
    team_norm == "AOTEAROANEWZEALAND"   ~ "NEWZEALAND",
    team_norm == "THEGAMBIA"            ~ "GAMBIA",
    team_norm == "TURKIYE"              ~ "TURKEY",
    team_norm == "TA14RKIYE"            ~ "TURKEY",   # encodage corrompu
    
    # ── Variantes de noms ────────────────────────────────────────────────────
    team_norm == "CABOVERDE"            ~ "CAPEVERDE",
    team_norm == "CAPEVERDEISLANDS"     ~ "CAPEVERDE",
    team_norm == "COTEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CATEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CONGODR"              ~ "DRCONGO",
    team_norm == "KOREAREPUBLIC"        ~ "SOUTHKOREA",
    team_norm == "KOREADPR"             ~ "NORTHKOREA",
    team_norm == "KYRGYZREPUBLIC"       ~ "KYRGYZSTAN",
    team_norm == "USA"                  ~ "UNITEDSTATES",
    team_norm == "USVIRGINISLANDS"      ~ "UNITEDSTATESVIRGINISLANDS",
    team_norm == "STKITTSANDNEVIS"      ~ "SAINTKITTSANDNEVIS",
    team_norm == "STLUCIA"              ~ "SAINTLUCIA",
    team_norm == "STVINCENTANDTHEGRENADINES" ~ "SAINTVINCENTANDTHEGRENADINES",
    team_norm == "BRUNEIDARUSSALAM"     ~ "BRUNEI",
    team_norm == "CHINESETAIPEI"        ~ "TAIWAN",
    team_norm == "IRIRAN"               ~ "IRAN",   # encodage corrompu
    
    # ── Variantes São Tomé ───────────────────────────────────────────────────
    team_norm == "SAOTOMEANDPRINCIPE"   ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMEEPRINCIPE"     ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMACANDPRANCIPE"  ~ "SAOTOMEETPRINCIPE",
    
    # ── Suffixes (unranked) → nom de base ────────────────────────────────────
    team_norm == "AMERICANSAMOAUNRANKED" ~ "AMERICANSAMOA",
    team_norm == "SAMOAUNRANKED"         ~ "SAMOA",
    team_norm == "TONGAUNRANKED"         ~ "TONGA",
    team_norm == "ERITREAUNRANKED"       ~ "ERITREA",
    team_norm == "ERITREA"               ~ "ERITREA",   # déjà bon mais absent de results
    
    # ── Typo ─────────────────────────────────────────────────────────────────
    team_norm == "CAMPOON"              ~ "CAMEROON",
    team_norm == "CURACAO"              ~ "CURACAO",   # déjà bon
    team_norm == "CURAAAO"              ~ "CURACAO",

    
    TRUE ~ team_norm
  ))
```

```{r}
results <- results |>
  mutate(
    home_norm = case_when(
      home_norm == "CURAAAO"             ~ "CURACAO",
      home_norm == "SAOTOMACANDPRANCIPE" ~ "SAOTOMEETPRINCIPE",
      TRUE ~ home_norm
    ),
    away_norm = case_when(
      away_norm == "CURAAAO"             ~ "CURACAO",
      away_norm == "SAOTOMACANDPRANCIPE" ~ "SAOTOMEETPRINCIPE",
      TRUE ~ away_norm
    )
  )
```

## Formatting the ranking

```{r}
ranking_clean <- ranking %>% 
  mutate(rank_points = total_points,
         date = as.Date(date)) %>% 
  filter(!is.na(rank_points)) %>% 
  select(date, team_norm, rank_points) %>% 
  mutate(team = team_norm) %>% 
  select(date, team, rank_points)
```

```{r}
top_140 <- ranking_clean %>% 
  filter(date == max(date, na.rm = TRUE)) %>% 
  slice_max(order_by = rank_points, n = 180, with_ties = FALSE) %>% 
  pull(team)
```

```{r}
ranking_worldcup <- ranking_clean %>% 
  filter(team %in% top_140)
```

```{r}
ranking_worldcup <- ranking_worldcup %>% 
  arrange(date, rank_points)
```

## Creation of the periods

```{r}
ranking_periods <- ranking_worldcup %>%
  arrange(date) %>%
  mutate(periods = dense_rank(date))

dates_periods <- ranking_periods %>%
  select(date, periods) %>%
  distinct()
```

```{r}
worldcup2026_periods <- results %>%
  full_join(dates_periods, by = c("date" = "date")) %>%
  arrange(date) %>%
  fill(periods, .direction = "down") %>% 
  filter(!is.na(home_team))
```

## Ranking ready to be used by footBayes package

```{r}
ranking_footBayes <- data.frame(
  team        = as.character(ranking_periods$team),
  rank_points = as.numeric(ranking_periods$rank_points),
  periods     = as.integer(ranking_periods$periods),
  stringsAsFactors = FALSE
)
```

## Formatting of results

```{r}
worldcup2026 <- worldcup2026_periods %>%
  mutate(date = as.Date(date)) %>%
  mutate(is_future = is.na(home_score) | is.na(away_score)) %>%
  arrange(is_future, date)

summary(worldcup2026)
table(worldcup2026$is_future)
```

```{r}
worldcup2026 <- worldcup2026 %>%
  filter((home_norm %in% ranking_footBayes$team) & (away_norm %in% ranking_footBayes$team) )
```

```{r}
friendly_teams <- worldcup2026 %>%
  filter(tournament == "Friendly") %>%
  select(home_norm, away_norm) %>%
  pivot_longer(
    cols = c(home_norm, away_norm),
    names_to = "side",
    values_to = "team"
  ) %>%
  distinct(team) %>%
  arrange(team)

wc48_teams <- worldcup2026 %>%
  filter(is_future == TRUE) %>%
  select(home_norm, away_norm)

wc48_teams <- unique(c(wc48_teams$home_norm, wc48_teams$away_norm))

length(wc48_teams)
```

```{r}
# The training data (the games played since the 1st of January of 2022)
played_matches <- worldcup2026 %>%
  filter(is_future == FALSE)

# Which games have missing values ?
played_matches_with_na <- played_matches %>%
  filter(if_any(everything(), is.na))

# How much games with missing values do we have ?
nrow(played_matches_with_na)
```

```{r}
# All the games with "suspicious" outcomes
worldcup2026_flagged <- worldcup2026 %>%
  mutate(
    total_goals = ifelse(is_future == FALSE, home_score + away_score, NA_real_),
    goal_diff   = ifelse(is_future == FALSE, abs(home_score - away_score), NA_real_),

    both_teams_in_wc48 = (home_norm %in% wc48_teams) &
      (away_norm %in% wc48_teams),

    friendly_suspect = (tournament == "Friendly") &
      (is_future == FALSE) &
      (
        goal_diff >= 5 |
          (total_goals >= 7 & goal_diff >= 4)
      ),

    remove_match = friendly_suspect & !both_teams_in_wc48
  )

# Friendly games with suspicious resultas 
friendly_suspects <- worldcup2026_flagged %>%
  filter(friendly_suspect == TRUE) %>%
  select(
    periods,
    date,
    home_team,
    home_norm,
    away_team,
    away_norm,
    home_score,
    away_score,
    tournament,
    total_goals,
    goal_diff,
    both_teams_in_wc48,
    remove_match,
    neutral
  ) %>%
  arrange(desc(goal_diff), desc(total_goals))

# Then we keep only the non suspicious games 
worldcup2026_clean <- worldcup2026_flagged %>%
  filter(remove_match == FALSE)
```

```{r}
cat("Matches before cleaning:", nrow(worldcup2026), "\n")
cat("Matches after cleaning:", nrow(worldcup2026_clean), "\n")
cat("Removed matches:", nrow(worldcup2026) - nrow(worldcup2026_clean), "\n")

table(worldcup2026$is_future)
table(worldcup2026_clean$is_future)

summary(worldcup2026_clean)
```

## Results ready for footBayes package

```{r}
results_footBayes <- worldcup2026_clean %>%
  arrange(is_future, date) %>%
  rename(
    home_goals = home_score,
    away_goals = away_score
  ) %>%
  select(
    periods,
    home_norm,
    away_norm,
    home_goals,
    away_goals,
    is_future
  ) %>% 
  mutate(home_team = home_norm,
         away_team = away_norm)
```

```{r}
n_future <- sum(results_footBayes$is_future)

cat("Number of future matches:", n_future, "\n")

if (n_future != 72) {
  warning("The number of future matches is not 72 after filtering.")
}

n_total <- nrow(results_footBayes)
n_train <- n_total - n_future

results_footBayes_ready <- results_footBayes %>% 
  select(periods, home_team, home_goals, away_team, away_goals)
```

# The predictions

## The Double Poisson model

```{r}
double_poisson <- stan_foot(
  data = results_footBayes_ready,
  model = "double_pois",
  predict = 72,
  ranking = ranking_footBayes,
  dynamic_type = "seasonal",
  home_effect = FALSE,
  chains = 4,
  iter_sampling = 1000,
  iter_warmup = 1000,
  seed = 123
)
```

```{r}
prob_double_poisson <- foot_prob(object = double_poisson, data = results_footBayes_ready)
```

```{r}
colnames(prob_double_poisson$prob_table) <- c(
      "home",
      "away",
      "home win",
      "draw",
      "away win",
      "mlo"
    )

prob_double_poisson$prob_table
```


## The Diaginal Inflated Bivariate Poisson model 