---
title: "TESTWC2026"
output: html_document
date: "2026-06-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### testing with 2026 


---
  title: "Predictions"
output: pdf_document
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Packages needed
library(tidyverse)
library(dplyr)
library(tidyr)
library(loo)
library(cmdstanr)
library(footBayes)
library(posterior)
library(bayesplot)
library(instantiate)
library(stringi)
```

# The theory

# The data

## Data loading and preprocessing

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("ranking_preprocessing.R")
```

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("ranking2025_11_19_preprocessing.R")
```

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("results_preprocessing.R")
```

```{r}
ranking <- rbind(ranking, ranking2025_11_19)
```


## All the teams are in the ranking ?

```{r}
all_rank_norms <- unique(ranking$team_norm)
res_teams <- unique(c(results$home_norm, results$away_norm))

unmatched_in_results <- setdiff(res_teams, all_rank_norms)

cat("\n=== Teams dans results sans correspondance dans ranking ===\n")
print(sort(unmatched_in_results))
```

## All the teams in ranking are in results 

```{r}
all_rank_norms <- unique(ranking$team_norm)
res_teams <- unique(c(results$home_norm, results$away_norm))

unmatched_in_results <- setdiff(res_teams, all_rank_norms)

cat("\n=== Teams dans results sans correspondance dans ranking ===\n")
print(sort(unmatched_in_results))
```

## Correction of the names which cause problems 

```{r}
ranking <- ranking |>
  mutate(team_norm = case_when(
    
    # ── Renommages officiels ──────────────────────────────────────────────────
    team_norm == "CZECHIA"              ~ "CZECHREPUBLIC",
    team_norm == "AOTEAROANEWZEALAND"   ~ "NEWZEALAND",
    team_norm == "THEGAMBIA"            ~ "GAMBIA",
    team_norm == "TURKIYE"              ~ "TURKEY",
    team_norm == "TA14RKIYE"            ~ "TURKEY",   # encodage corrompu
    
    # ── Variantes de noms ────────────────────────────────────────────────────
    team_norm == "CABOVERDE"            ~ "CAPEVERDE",
    team_norm == "CAPEVERDEISLANDS"     ~ "CAPEVERDE",
    team_norm == "COTEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CATEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CONGODR"              ~ "DRCONGO",
    team_norm == "KOREAREPUBLIC"        ~ "SOUTHKOREA",
    team_norm == "KOREADPR"             ~ "NORTHKOREA",
    team_norm == "KYRGYZREPUBLIC"       ~ "KYRGYZSTAN",
    team_norm == "USA"                  ~ "UNITEDSTATES",
    team_norm == "USVIRGINISLANDS"      ~ "UNITEDSTATESVIRGINISLANDS",
    team_norm == "STKITTSANDNEVIS"      ~ "SAINTKITTSANDNEVIS",
    team_norm == "STLUCIA"              ~ "SAINTLUCIA",
    team_norm == "STVINCENTANDTHEGRENADINES" ~ "SAINTVINCENTANDTHEGRENADINES",
    team_norm == "BRUNEIDARUSSALAM"     ~ "BRUNEI",
    team_norm == "CHINESETAIPEI"        ~ "TAIWAN",
    team_norm == "IRIRAN"               ~ "IRAN",   # encodage corrompu
    
    # ── Variantes São Tomé ───────────────────────────────────────────────────
    team_norm == "SAOTOMEANDPRINCIPE"   ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMEEPRINCIPE"     ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMACANDPRANCIPE"  ~ "SAOTOMEETPRINCIPE",
    
    # ── Suffixes (unranked) → nom de base ────────────────────────────────────
    team_norm == "AMERICANSAMOAUNRANKED" ~ "AMERICANSAMOA",
    team_norm == "SAMOAUNRANKED"         ~ "SAMOA",
    team_norm == "TONGAUNRANKED"         ~ "TONGA",
    team_norm == "ERITREAUNRANKED"       ~ "ERITREA",
    team_norm == "ERITREA"               ~ "ERITREA",   # déjà bon mais absent de results
    
    # ── Typo ─────────────────────────────────────────────────────────────────
    team_norm == "CAMPOON"              ~ "CAMEROON",
    team_norm == "CURACAO"              ~ "CURACAO",   # déjà bon
    team_norm == "CURAAAO"              ~ "CURACAO",
    
    
    TRUE ~ team_norm
  ))
```

```{r}
results <- results |>
  mutate(
    home_norm = case_when(
      home_norm == "CURAAAO"             ~ "CURACAO",
      home_norm == "SAOTOMACANDPRANCIPE" ~ "SAOTOMEETPRINCIPE",
      TRUE ~ home_norm
    ),
    away_norm = case_when(
      away_norm == "CURAAAO"             ~ "CURACAO",
      away_norm == "SAOTOMACANDPRANCIPE" ~ "SAOTOMEETPRINCIPE",
      TRUE ~ away_norm
    )
  )
```

## Formatting the ranking

```{r}
ranking_clean <- ranking %>% 
  mutate(rank_points = total_points,
         date = as.Date(date)) %>% 
  filter(!is.na(rank_points)) %>% 
  select(date, team_norm, rank_points) %>% 
  mutate(team = team_norm) %>% 
  select(date, team, rank_points)
```

```{r}
top_140 <- ranking_clean %>% 
  filter(date == max(date, na.rm = TRUE)) %>% 
  slice_max(order_by = rank_points, n = 180, with_ties = FALSE) %>% 
  pull(team)
```

```{r}
ranking_worldcup <- ranking_clean %>% 
  filter(team %in% top_140)
```

```{r}
ranking_worldcup <- ranking_worldcup %>% 
  arrange(date, rank_points)
```

## Creation of the periods

```{r}
ranking_periods <- ranking_worldcup %>%
  arrange(date) %>%
  mutate(periods = dense_rank(date))

dates_periods <- ranking_periods %>%
  select(date, periods) %>%
  distinct()
```

```{r}
worldcup2026_periods <- results %>%
  full_join(dates_periods, by = c("date" = "date")) %>%
  arrange(date) %>%
  fill(periods, .direction = "down") %>% 
  filter(!is.na(home_team))
```

## Ranking ready to be used by footBayes package

```{r}
ranking_footBayes <- data.frame(
  team        = as.character(ranking_periods$team),
  rank_points = as.numeric(ranking_periods$rank_points),
  periods     = as.integer(ranking_periods$periods),
  stringsAsFactors = FALSE
)
```

## Formatting of results

```{r}
worldcup2026 <- worldcup2026_periods %>%
  mutate(date = as.Date(date)) %>%
  mutate(is_future = is.na(home_score) | is.na(away_score)) %>%
  arrange(is_future, date)

summary(worldcup2026)
table(worldcup2026$is_future)
```

```{r}
worldcup2026 <- worldcup2026 %>%
  filter((home_norm %in% ranking_footBayes$team) & (away_norm %in% ranking_footBayes$team) )
```

```{r}
friendly_teams <- worldcup2026 %>%
  filter(tournament == "Friendly") %>%
  select(home_norm, away_norm) %>%
  pivot_longer(
    cols = c(home_norm, away_norm),
    names_to = "side",
    values_to = "team"
  ) %>%
  distinct(team) %>%
  arrange(team)

wc48_teams <- worldcup2026 %>%
  filter(is_future == TRUE) %>%
  select(home_norm, away_norm)

wc48_teams <- unique(c(wc48_teams$home_norm, wc48_teams$away_norm))

length(wc48_teams)
```

```{r}
# The training data (the games played since the 1st of January of 2022)
played_matches <- worldcup2026 %>%
  filter(is_future == FALSE)

# Which games have missing values ?
played_matches_with_na <- played_matches %>%
  filter(if_any(everything(), is.na))

# How much games with missing values do we have ?
nrow(played_matches_with_na)
```

```{r}
# All the games with "suspicious" outcomes
worldcup2026_flagged <- worldcup2026 %>%
  mutate(
    total_goals = ifelse(is_future == FALSE, home_score + away_score, NA_real_),
    goal_diff   = ifelse(is_future == FALSE, abs(home_score - away_score), NA_real_),
    
    both_teams_in_wc48 = (home_norm %in% wc48_teams) &
      (away_norm %in% wc48_teams),
    
    friendly_suspect = (tournament == "Friendly") &
      (is_future == FALSE) &
      (
        goal_diff >= 5 |
          (total_goals >= 7 & goal_diff >= 4)
      ),
    
    remove_match = friendly_suspect & !both_teams_in_wc48
  )

# Friendly games with suspicious resultas 
friendly_suspects <- worldcup2026_flagged %>%
  filter(friendly_suspect == TRUE) %>%
  select(
    periods,
    date,
    home_team,
    home_norm,
    away_team,
    away_norm,
    home_score,
    away_score,
    tournament,
    total_goals,
    goal_diff,
    both_teams_in_wc48,
    remove_match,
    neutral
  ) %>%
  arrange(desc(goal_diff), desc(total_goals))

# Then we keep only the non suspicious games 
worldcup2026_clean <- worldcup2026_flagged %>%
  filter(remove_match == FALSE)
```

```{r}
cat("Matches before cleaning:", nrow(worldcup2026), "\n")
cat("Matches after cleaning:", nrow(worldcup2026_clean), "\n")
cat("Removed matches:", nrow(worldcup2026) - nrow(worldcup2026_clean), "\n")

table(worldcup2026$is_future)
table(worldcup2026_clean$is_future)

summary(worldcup2026_clean)
```

## Results ready for footBayes package

```{r}
results_footBayes <- worldcup2026_clean %>%
  arrange(is_future, date) %>%
  rename(
    home_goals = home_score,
    away_goals = away_score
  ) %>%
  select(
    periods,
    home_norm,
    away_norm,
    home_goals,
    away_goals,
    is_future
  ) %>% 
  mutate(home_team = home_norm,
         away_team = away_norm)
```

```{r}
n_future <- sum(results_footBayes$is_future)

cat("Number of future matches:", n_future, "\n")

if (n_future != 72) {
  warning("The number of future matches is not 72 after filtering.")
}

n_total <- nrow(results_footBayes)
n_train <- n_total - n_future

results_footBayes_ready <- results_footBayes %>% 
  select(periods, home_team, home_goals, away_team, away_goals)
```




```{r}
# Out-of-sample validation on period 29
# Diagonal-inflated bivariate Poisson model


# - Use all played matches from periods 1 to 28 as training.
# - In period 29, keep the first 12 matches as training.
# - In period 29, hide the remaining 18 matches and use them as test matches.
# - Future World Cup matches are removed from this validation test 

period_to_test <- 29
n_train_period_29 <- 12

# Keep only played matches with valid ranking periods
worldcup_validation_p29 <- worldcup2026_clean %>%
  filter(is_future == FALSE) %>%
  filter(!is.na(periods)) %>%
  arrange(periods, date)

# Check the number of played matches in period 29.
n_period_29 <- worldcup_validation_p29 %>%
  filter(periods == period_to_test) %>%
  nrow()

cat("Number of played matches in period 29:", n_period_29, "\n")
cat("Number of period 29 matches used for training:", n_train_period_29, "\n")
cat("Number of period 29 matches used for test:", n_period_29 - n_train_period_29, "\n")

# Stop if period 29 has too few matches
if (n_period_29 <= n_train_period_29) {
  stop("Period 29 does not have enough matches for this validation split.")
}

# Create the validation split.
worldcup_validation_p29 <- worldcup_validation_p29 %>%
  group_by(periods) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(
    row_in_period = row_number(),
    is_test = periods == period_to_test &
      row_in_period > n_train_period_29,
    
    # Hide the scores of the test matches.
    home_score_model = ifelse(is_test, NA_real_, home_score),
    away_score_model = ifelse(is_test, NA_real_, away_score)
  ) %>%
  ungroup() %>%
  arrange(is_test, date)

# Count test matches
n_test_p29 <- sum(worldcup_validation_p29$is_test)

cat("Number of validation test matches:", n_test_p29, "\n")

# Save the true results of the hidden test matches
true_validation_p29 <- worldcup_validation_p29 %>%
  filter(is_test == TRUE) %>%
  mutate(
    real_result = case_when(
      home_score > away_score ~ "home win",
      home_score == away_score ~ "draw",
      home_score < away_score ~ "away win"
    )
  ) %>%
  select(
    date,
    periods,
    home_norm,
    away_norm,
    home_score,
    away_score,
    real_result
  )

# Prepare data for footBayes

# The test matches are at the end because footBayes predicts the last n matches.
results_footBayes_validation_p29_ready <- worldcup_validation_p29 %>%
  transmute(
    periods = as.integer(periods),
    home_team = as.character(home_norm),
    home_goals = as.numeric(home_score_model),
    away_team = as.character(away_norm),
    away_goals = as.numeric(away_score_model)
  )

# Keep the ranking periods used in this validation dataset.
used_periods_p29 <- sort(unique(results_footBayes_validation_p29_ready$periods))

ranking_footBayes_validation_p29 <- ranking_footBayes %>%
  filter(periods %in% used_periods_p29) %>%
  transmute(
    team = as.character(team),
    rank_points = as.numeric(rank_points),
    periods = as.integer(periods)
  )

# check before MCMC

cat("NA periods in validation data:", sum(is.na(results_footBayes_validation_p29_ready$periods)), "\n")
cat("NA periods in validation ranking:", sum(is.na(ranking_footBayes_validation_p29$periods)), "\n")

training_periods_p29 <- results_footBayes_validation_p29_ready %>%
  filter(!is.na(home_goals), !is.na(away_goals)) %>%
  pull(periods) %>%
  unique() %>%
  sort()

ranking_periods_p29 <- ranking_footBayes_validation_p29 %>%
  pull(periods) %>%
  unique() %>%
  sort()

cat("Training periods:\n")
print(training_periods_p29)

cat("Ranking periods:\n")
print(ranking_periods_p29)

cat("Periods in ranking but not in training:\n")
print(setdiff(ranking_periods_p29, training_periods_p29))

cat("Periods in training but not in ranking:\n")
print(setdiff(training_periods_p29, ranking_periods_p29))

# Stop if periods are not aligned
if (!identical(training_periods_p29, ranking_periods_p29)) {
  stop("Training periods and ranking periods are not aligned.")
}

# Check that the number of prediction rows is correct.
n_missing_scores_p29 <- sum(
  is.na(results_footBayes_validation_p29_ready$home_goals) |
    is.na(results_footBayes_validation_p29_ready$away_goals)
)

cat("Number of rows with hidden scores:", n_missing_scores_p29, "\n")

if (n_missing_scores_p29 != n_test_p29) {
  stop("The number of hidden-score rows does not match n_test_p29.")
}

# ====================================
# MCMC model for validation
# Diagonal-inflated bivariate Poisson


diag_infl_biv_pois_validation_p29 <- stan_foot(
  data = results_footBayes_validation_p29_ready,
  model = "diag_infl_biv_pois",
  predict = n_test_p29,
  ranking = ranking_footBayes_validation_p29,
  dynamic_type = "seasonal",
  home_effect = FALSE,
  chains = 4,
  iter_sampling = 1000,
  iter_warmup = 1000,
  seed = 123,
  init = 0
)


# Predicted probabilities


prob_validation_p29 <- foot_prob(
  object = diag_infl_biv_pois_validation_p29,
  data = results_footBayes_validation_p29_ready
)

colnames(prob_validation_p29$prob_table) <- c(
  "home",
  "away",
  "home win",
  "draw",
  "away win",
  "mlo"
)

pred_validation_p29 <- prob_validation_p29$prob_table

# Combine true results and predicted probabilities.
evaluation_p29 <- bind_cols(
  true_validation_p29,
  pred_validation_p29
)

# Evaluation metrics

# Accuracy
accuracy_p29 <- mean(evaluation_p29$mlo == evaluation_p29$real_result)

# Brier Score: measures the quality of the predicted probabilities
evaluation_p29 <- evaluation_p29 %>%
  mutate(
    y_home = ifelse(real_result == "home win", 1, 0),
    y_draw = ifelse(real_result == "draw", 1, 0),
    y_away = ifelse(real_result == "away win", 1, 0),
    
    brier = (`home win` - y_home)^2 +
      (`draw` - y_draw)^2 +
      (`away win` - y_away)^2
  )

brier_p29 <- mean(evaluation_p29$brier)

# Log-loss: penalizes confident wrong predictions strongly
epsilon <- 1e-15

evaluation_p29 <- evaluation_p29 %>%
  mutate(
    prob_real = case_when(
      real_result == "home win" ~ `home win`,
      real_result == "draw" ~ `draw`,
      real_result == "away win" ~ `away win`
    ),
    logloss = -log(pmax(prob_real, epsilon))
  )

logloss_p29 <- mean(evaluation_p29$logloss)

# results.
cat("Period 29 validation accuracy:", accuracy_p29, "\n")
cat("Period 29 validation Brier Score:", brier_p29, "\n")
cat("Period 29 validation Log-loss:", logloss_p29, "\n")

evaluation_p29



```



































