---
title: "Predictions after groupstages"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Packages needed
library(dplyr)
library(tidyr)
library(loo)
library(cmdstanr)
library(footBayes)
library(posterior)
library(bayesplot)
library(instantiate)
library(stringi)
library(stringr)
library(readr)
```

# The data for the model which allow us to have posteriors (future priors)

## Data loading and preprocessing

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("~/work/World-Cup-2026/codes/ranking_preprocessing.R")
```

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("~/work/World-Cup-2026/codes/ranking2025_11_19_preprocessing.R")
```

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("~/work/World-Cup-2026/codes/results_preprocessing.R")
```

```{r}
ranking <- rbind(ranking, ranking2025_11_19)
```


## All the teams are in the ranking ?

```{r}
all_rank_norms <- unique(ranking$team_norm)
res_teams <- unique(c(results$home_norm, results$away_norm))

unmatched_in_results <- setdiff(res_teams, all_rank_norms)

cat("\n=== Teams dans results sans correspondance dans ranking ===\n")
print(sort(unmatched_in_results))
```

## All the teams in ranking are in results 

```{r}
all_rank_norms <- unique(ranking$team_norm)
res_teams <- unique(c(results$home_norm, results$away_norm))

unmatched_in_results <- setdiff(res_teams, all_rank_norms)

cat("\n=== Teams dans results sans correspondance dans ranking ===\n")
print(sort(unmatched_in_results))
```

## Correction of the names which cause problems 

```{r}
ranking <- ranking |>
  mutate(team_norm = case_when(
    
    # ── Renommages officiels ──────────────────────────────────────────────────
    team_norm == "CZECHIA"              ~ "CZECHREPUBLIC",
    team_norm == "AOTEAROANEWZEALAND"   ~ "NEWZEALAND",
    team_norm == "THEGAMBIA"            ~ "GAMBIA",
    team_norm == "TURKIYE"              ~ "TURKEY",
    team_norm == "TA14RKIYE"            ~ "TURKEY",   # encodage corrompu
    
    # ── Variantes de noms ────────────────────────────────────────────────────
    team_norm == "CABOVERDE"            ~ "CAPEVERDE",
    team_norm == "CAPEVERDEISLANDS"     ~ "CAPEVERDE",
    team_norm == "COTEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CATEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CONGODR"              ~ "DRCONGO",
    team_norm == "KOREAREPUBLIC"        ~ "SOUTHKOREA",
    team_norm == "KOREADPR"             ~ "NORTHKOREA",
    team_norm == "KYRGYZREPUBLIC"       ~ "KYRGYZSTAN",
    team_norm == "USA"                  ~ "UNITEDSTATES",
    team_norm == "USVIRGINISLANDS"      ~ "UNITEDSTATESVIRGINISLANDS",
    team_norm == "STKITTSANDNEVIS"      ~ "SAINTKITTSANDNEVIS",
    team_norm == "STLUCIA"              ~ "SAINTLUCIA",
    team_norm == "STVINCENTANDTHEGRENADINES" ~ "SAINTVINCENTANDTHEGRENADINES",
    team_norm == "BRUNEIDARUSSALAM"     ~ "BRUNEI",
    team_norm == "CHINESETAIPEI"        ~ "TAIWAN",
    team_norm == "IRIRAN"               ~ "IRAN",   # encodage corrompu
    
    # ── Variantes São Tomé ───────────────────────────────────────────────────
    team_norm == "SAOTOMEANDPRINCIPE"   ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMEEPRINCIPE"     ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMACANDPRANCIPE"  ~ "SAOTOMEETPRINCIPE",
    
    # ── Suffixes (unranked) → nom de base ────────────────────────────────────
    team_norm == "AMERICANSAMOAUNRANKED" ~ "AMERICANSAMOA",
    team_norm == "SAMOAUNRANKED"         ~ "SAMOA",
    team_norm == "TONGAUNRANKED"         ~ "TONGA",
    team_norm == "ERITREAUNRANKED"       ~ "ERITREA",
    team_norm == "ERITREA"               ~ "ERITREA",   # déjà bon mais absent de results
    
    # ── Typo ─────────────────────────────────────────────────────────────────
    team_norm == "CAMPOON"              ~ "CAMEROON",
    team_norm == "CURACAO"              ~ "CURACAO",   # déjà bon
    team_norm == "CURAAAO"              ~ "CURACAO",

    
    TRUE ~ team_norm
  ))
```

```{r}
results <- results |>
  mutate(
    home_norm = case_when(
      home_norm == "CURAAAO"             ~ "CURACAO",
      home_norm == "SAOTOMACANDPRANCIPE" ~ "SAOTOMEETPRINCIPE",
      TRUE ~ home_norm
    ),
    away_norm = case_when(
      away_norm == "CURAAAO"             ~ "CURACAO",
      away_norm == "SAOTOMACANDPRANCIPE" ~ "SAOTOMEETPRINCIPE",
      TRUE ~ away_norm
    )
  )
```

## Formatting the ranking

```{r}
ranking_clean <- ranking %>% 
  mutate(rank_points = total_points,
         date = as.Date(date)) %>% 
  filter(!is.na(rank_points)) %>% 
  select(date, team_norm, rank_points) %>% 
  mutate(team = team_norm) %>% 
  select(date, team, rank_points)
```

```{r}
top_140 <- ranking_clean %>% 
  filter(date == max(date, na.rm = TRUE)) %>% 
  slice_max(order_by = rank_points, n = 180, with_ties = FALSE) %>% 
  pull(team)
```

```{r}
ranking_worldcup <- ranking_clean %>% 
  filter(team %in% top_140)
```

```{r}
ranking_worldcup <- ranking_worldcup %>% 
  arrange(date, rank_points)
```

## Creation of the periods

```{r}
ranking_periods <- ranking_worldcup %>%
  arrange(date) %>%
  mutate(periods = dense_rank(date))

dates_periods <- ranking_periods %>%
  select(date, periods) %>%
  distinct()
```

```{r}
worldcup2026_periods <- results %>%
  full_join(dates_periods, by = c("date" = "date")) %>%
  arrange(date) %>%
  fill(periods, .direction = "down") %>% 
  filter(!is.na(home_team))
```

## Ranking ready to be used by footBayes package

```{r}
ranking_footBayes <- data.frame(
  team        = as.character(ranking_periods$team),
  rank_points = as.numeric(ranking_periods$rank_points),
  periods     = as.integer(ranking_periods$periods),
  stringsAsFactors = FALSE
)
```

## Formatting of results

```{r}
worldcup2026 <- worldcup2026_periods %>%
  mutate(date = as.Date(date)) %>%
  mutate(is_future = is.na(home_score) | is.na(away_score)) %>%
  arrange(is_future, date)

summary(worldcup2026)
table(worldcup2026$is_future)
```

```{r}
worldcup2026 <- worldcup2026 %>%
  filter((home_norm %in% ranking_footBayes$team) & (away_norm %in% ranking_footBayes$team) )
```

```{r}
friendly_teams <- worldcup2026 %>%
  filter(tournament == "Friendly") %>%
  select(home_norm, away_norm) %>%
  pivot_longer(
    cols = c(home_norm, away_norm),
    names_to = "side",
    values_to = "team"
  ) %>%
  distinct(team) %>%
  arrange(team)

wc48_teams <- worldcup2026 %>%
  filter(is_future == TRUE) %>%
  select(home_norm, away_norm)

wc48_teams <- unique(c(wc48_teams$home_norm, wc48_teams$away_norm))

length(wc48_teams)
```

```{r}
# The training data (the games played since the 1st of January of 2022)
played_matches <- worldcup2026 %>%
  filter(is_future == FALSE)

# Which games have missing values ?
played_matches_with_na <- played_matches %>%
  filter(if_any(everything(), is.na))

# How much games with missing values do we have ?
nrow(played_matches_with_na)
```

```{r}
# All the games with "suspicious" outcomes
worldcup2026_flagged <- worldcup2026 %>%
  mutate(
    total_goals = ifelse(is_future == FALSE, home_score + away_score, NA_real_),
    goal_diff   = ifelse(is_future == FALSE, abs(home_score - away_score), NA_real_),

    both_teams_in_wc48 = (home_norm %in% wc48_teams) &
      (away_norm %in% wc48_teams),

    friendly_suspect = (tournament == "Friendly") &
      (is_future == FALSE) &
      (
        goal_diff >= 5 |
          (total_goals >= 7 & goal_diff >= 4)
      ),

    remove_match = friendly_suspect & !both_teams_in_wc48
  )

# Friendly games with suspicious resultas 
friendly_suspects <- worldcup2026_flagged %>%
  filter(friendly_suspect == TRUE) %>%
  select(
    periods,
    date,
    home_team,
    home_norm,
    away_team,
    away_norm,
    home_score,
    away_score,
    tournament,
    total_goals,
    goal_diff,
    both_teams_in_wc48,
    remove_match,
    neutral
  ) %>%
  arrange(desc(goal_diff), desc(total_goals))

# Then we keep only the non suspicious games 
worldcup2026_clean <- worldcup2026_flagged %>%
  filter(remove_match == FALSE)
```

```{r}
cat("Matches before cleaning:", nrow(worldcup2026), "\n")
cat("Matches after cleaning:", nrow(worldcup2026_clean), "\n")
cat("Removed matches:", nrow(worldcup2026) - nrow(worldcup2026_clean), "\n")

table(worldcup2026$is_future)
table(worldcup2026_clean$is_future)

summary(worldcup2026_clean)
```

## Results ready for footBayes package

```{r}
results_footBayes <- worldcup2026_clean %>%
  arrange(is_future, date) %>%
  rename(
    home_goals = home_score,
    away_goals = away_score
  ) %>%
  select(
    periods,
    home_norm,
    away_norm,
    home_goals,
    away_goals,
    is_future
  ) %>% 
  mutate(home_team = home_norm,
         away_team = away_norm)
```

```{r}
n_future <- sum(results_footBayes$is_future)

cat("Number of future matches:", n_future, "\n")

if (n_future != 72) {
  warning("The number of future matches is not 72 after filtering.")
}

n_total <- nrow(results_footBayes)
n_train <- n_total - n_future

results_footBayes_ready <- results_footBayes %>% 
  select(periods, home_team, home_goals, away_team, away_goals)
```



```{r}
groupstage1 <- results_footBayes_ready %>% 
  filter(!is.na(home_goals))

teams <- unique(c(groupstage1$home_team, groupstage1$away_team))

ranking_footBayes_clean <- ranking_footBayes %>%
  filter(team %in% teams)
```

# The model

```{r}
diag_infl_without_world_cup <- stan_foot(
 data = groupstage1,
 model = "diag_infl_biv_pois",
 ranking = ranking_footBayes_clean,
 dynamic_type = "seasonal",
 init = 0,
 home_effect = FALSE,
 chains = 4, parallel_chains = 4,
 iter_sampling = 1000,
 iter_warmup = 2000,        
 adapt_delta = 0.99,        
 max_treedepth = 14, 
 seed = 123
)
```

```{r}
# saveRDS(diag_infl_without_world_cup, file = "~/work/World-Cup-2026/results/diag_infl_without_world_cup.rds")
```

```{r}
diag_infl_without_world_cup <- readRDS(file = "~/work/World-Cup-2026/results/diag_infl_without_world_cup.rds")
```

# Extraction of the posteriors and initilization of the priors

```{r}
teams_m1   <- unique(groupstage1$home_team)
n_teams_m1 <- length(teams_m1)
n_chains <- 4
```

```{r}
draws_df <- diag_infl_without_world_cup$fit$draws(format = "df")
```

```{r}
att_cols <- grep("^att_raw_centered\\[", colnames(draws_df), value = TRUE)
def_cols <- grep("^def_raw_centered\\[", colnames(draws_df), value = TRUE)
```

```{r}
extract_ij <- function(cols) {
  idx_str <- regmatches(cols, regexpr("\\d+,\\d+", cols))
  i <- as.integer(sub(",.*", "", idx_str))
  j <- as.integer(gsub("\\]", "", sub(".*,", "", idx_str)))
  list(i = i, j = j, nrow = max(i), ncol = max(j))
}
```

```{r}
ij_att <- extract_ij(att_cols)
ij_def <- extract_ij(def_cols)

n_periods_m1  <- ij_att$nrow  
n_teams_check <- ij_att$ncol  
```

```{r}
last_period <- n_periods_m1

missing_teams <- setdiff(wc48_teams, teams_m1)
if (length(missing_teams) > 0) {
  warning("Équipes de wc48_teams absentes de teams_m1 (groupstage1) : ",
          paste(missing_teams, collapse = ", "))
}

keep_team_idx <- match(wc48_teams, teams_m1)
keep_team_idx <- keep_team_idx[!is.na(keep_team_idx)]  
teams_kept    <- teams_m1[keep_team_idx] 
n_teams_kept <- length(keep_team_idx)

att_mask <- (ij_att$i == last_period) & (ij_att$j %in% keep_team_idx)
def_mask <- (ij_def$i == last_period) & (ij_def$j %in% keep_team_idx)

att_cols_last <- att_cols[att_mask]
def_cols_last <- def_cols[def_mask]
att_j_last    <- ij_att$j[att_mask]
def_j_last    <- ij_def$j[def_mask]
```

```{r}
pos_in_kept <- match(att_j_last, keep_team_idx)   

parse_vector_kept <- function(cols, pos_idx, n_teams_kept, draws_row) {
  vec <- numeric(n_teams_kept)
  for (k in seq_along(cols)) vec[pos_idx[k]] <- as.numeric(draws_row[[cols[k]]])
  vec
}

pos_in_kept_def <- match(def_j_last, keep_team_idx)
```

```{r}
inits_pred <- lapply(1:n_chains, function(c) {
  draw_idx  <- sample(nrow(draws_df), 1)
  draws_row <- draws_df[draw_idx, ]

  att_last <- parse_vector_kept(att_cols_last, pos_in_kept, n_teams_kept, draws_row)
  def_last <- parse_vector_kept(def_cols_last, pos_in_kept_def, n_teams_kept, draws_row)

  list(
    att_raw = att_last, 
    def_raw = def_last,
    rho           = as.numeric(draws_row[["rho"]]),
    gamma         = as.numeric(draws_row[["gamma"]]),
    sigma_att     = as.numeric(draws_row[["sigma_att[1]"]]),
    sigma_def     = as.numeric(draws_row[["sigma_def[1]"]]),
    prob_of_draws = as.numeric(draws_row[["prob_of_draws"]])
  )
})
```


# The data for the games we want to predict

## The results

```{r}
results_groupstage1 <- data.frame(
  periods    = rep(31, 24),   
  home_team  = c("MEXICO","SOUTHKOREA","CANADA","UNITEDSTATES",
                 "QATAR","BRAZIL","HAITI","AUSTRALIA",
                 "GERMANY", "NETHERLANDS","IVORYCOAST","SWEDEN",
                 "SPAIN","BELGIUM","SAUDIARABIA","IRAN",
                 "FRANCE","IRAQ","ARGENTINA","AUSTRIA",
                 "PORTUGAL", "ENGLAND", "GHANA", "UZBEKISTAN"),
  home_goals = c(2,2,1,4,1,1,0,2,
                 5,2,1,5,0,1,1,2,
                 3,0,3,3,1,4,1,1), 
  away_team  = c("SOUTHAFRICA","CZECHREPUBLIC","BOSNIAANDHERZEGOVINA","PARAGUAY",
                 "SWITZERLAND","MOROCCO","SCOTLAND","TURKEY",
                 "CURACAO", "JAPAN","ECUADOR","TUNISIA",
                 "CAPEVERDE","EGYPT","URUGUAY","NEWZEALAND",
                 "SENEGAL","NORWAY","ALGERIA","JORDAN",
                 "DRCONGO", "CROATIA", "PANAMA","COLOMBIA"),
  away_goals = c(0,1,1,1,1,1,1,0,
                 1,2,0,1,0,1,1,2,
                 1,2,0,1,1,2,0,3)
)
```

```{r}
results_groupstage2 <- data.frame(
  periods    = rep(31, 24),   
  home_team  = c("CZECHREPUBLIC","SWITZERLAND","CANADA","MEXICO",
                 "UNITEDSTATES","SCOTLAND","BRAZIL","TURKEY",
                 "NETHERLANDS","GERMANY","ECUADOR","TUNISIA",
                 "SPAIN","BELGIUM","URUGUAY","NEWZEALAND",
                 "ARGENTINA","FRANCE","NORWAY","JORDAN",
                 "PORTUGAL","ENGLAND","PANAMA","COLOMBIA"),
  home_goals = c(1,4,4,1,2,0,3,0,
                 5,2,0,0,4,0,2,1,
                 2,3,3,1,4,0,0,1), 
  away_team  = c("SOUTHAFRICA","BOSNIAANDHERZEGOVINA","QATAR","SOUTHKOREA",
                 "AUSTRALIA","MOROCCO","HAITI","PARAGUAY",
                 "SWEDEN","IVORYCOAST","CURACAO","JAPAN",
                 "SAUDIARABIA","IRAN","CAPEVERDE","EGYPT",
                 "AUSTRIA","IRAQ","SENEGAL","ALGERIA",
                 "UZBEKISTAN","GHANA","CROATIA","DRCONGO"),
  away_goals = c(1,1,0,0,0,1,0,1,
                 1,1,0,4,0,0,2,3,
                 0,0,2,2,0,0,1,0)
)
```

```{r}
results_groupstage3 <- data.frame(
  periods = rep(31,12),
  home_team = c("SWITZERLAND","BOSNIAANDHERZEGOVINA","MOROCCO","SCOTLAND",
                 "SOUTHAFRICA","CZECHREPUBLIC","CURACAO","ECUADOR",
                 "TUNISIA","JAPAN","TURKEY","PARAGUAY",
                "NORWAY","SENEGAL","CAPEVERDE","URUGUAY",
                "NEWZEALAND","EGYPT","PANAMA","CROATIA",
                "COLOMBIA","DRCONGO","ALGERIA","JORDAN"),
  home_goals = c(2,3,4,0,1,0,0,2,
                 1,1,3,0,1,5,0,0,
                 1,1,0,2,0,3,3,1),
  away_team = c("CANADA","QATAR","HAITI","BRAZIL",
                 "SOUTHKOREA","MEXICO","IVORYCOAST","GERMANY",
                 "NETHERLANDS","SWEDEN","UNITEDSTATES","AUSTRALIA",
                "FRANCE","IRAQ","SAUDIARABIA","SPAIN",
                "BELGIUM","IRAN","ENGLAND","GHANA",
                "PORTUGAL","UZBEKISTAN","AUSTRIA","ARGENTINA"),
  away_goals = c(1,1,2,3,0,3,2,1,
                 3,1,2,0,4,0,0,1,
                 5,1,2,1,0,1,3,3)
)
```

```{r}
next_games <- data.frame(
  periods = rep(31,16),
  home_team = c("SOUTHAFRICA","BRAZIL","GERMANY","NETHERLANDS",
              "IVORYCOAST","FRANCE","MEXICO","ENGLAND",
              "BELGIUM","UNITEDSTATES","SPAIN","PORTUGAL",
              "SWITZERLAND","AUSTRALIA","ARGENTINA","COLOMBIA"),
  home_goals = rep(NA,16),
  away_team = c("CANADA","JAPAN","PARAGUAY","MOROCCO",
                "NORWAY","SWEDEN","ECUADOR","DRCONGO",
                "SENEGAL","BOSNIAANDHERZEGOVINA","AUSTRIA","CROATIA",
                "ALGERIA","EGYPT","CAPEVERDE","GHANA"),
  away_goals = rep(NA,16)
)
```

```{r}
final_data <- rbind(results_groupstage1, results_groupstage2, results_groupstage3, next_games)
```

## The ranking

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("~/work/World-Cup-2026/codes/ranking2026_06_11_preprocessing.R")
ranking <- ranking2026_06_11
```

## Correction of the names which cause problems 

```{r}
ranking <- ranking |>
  mutate(team_norm = case_when(
    
    # ── Renommages officiels ──────────────────────────────────────────────────
    team_norm == "CZECHIA"              ~ "CZECHREPUBLIC",
    team_norm == "AOTEAROANEWZEALAND"   ~ "NEWZEALAND",
    team_norm == "THEGAMBIA"            ~ "GAMBIA",
    team_norm == "TURKIYE"              ~ "TURKEY",
    team_norm == "TA14RKIYE"            ~ "TURKEY",   # encodage corrompu
    
    # ── Variantes de noms ────────────────────────────────────────────────────
    team_norm == "CABOVERDE"            ~ "CAPEVERDE",
    team_norm == "CAPEVERDEISLANDS"     ~ "CAPEVERDE",
    team_norm == "COTEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CATEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CONGODR"              ~ "DRCONGO",
    team_norm == "KOREAREPUBLIC"        ~ "SOUTHKOREA",
    team_norm == "KOREADPR"             ~ "NORTHKOREA",
    team_norm == "KYRGYZREPUBLIC"       ~ "KYRGYZSTAN",
    team_norm == "USA"                  ~ "UNITEDSTATES",
    team_norm == "USVIRGINISLANDS"      ~ "UNITEDSTATESVIRGINISLANDS",
    team_norm == "STKITTSANDNEVIS"      ~ "SAINTKITTSANDNEVIS",
    team_norm == "STLUCIA"              ~ "SAINTLUCIA",
    team_norm == "STVINCENTANDTHEGRENADINES" ~ "SAINTVINCENTANDTHEGRENADINES",
    team_norm == "BRUNEIDARUSSALAM"     ~ "BRUNEI",
    team_norm == "CHINESETAIPEI"        ~ "TAIWAN",
    team_norm == "IRIRAN"               ~ "IRAN",   # encodage corrompu
    
    # ── Variantes São Tomé ───────────────────────────────────────────────────
    team_norm == "SAOTOMEANDPRINCIPE"   ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMEEPRINCIPE"     ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMACANDPRANCIPE"  ~ "SAOTOMEETPRINCIPE",
    
    # ── Suffixes (unranked) → nom de base ────────────────────────────────────
    team_norm == "AMERICANSAMOAUNRANKED" ~ "AMERICANSAMOA",
    team_norm == "SAMOAUNRANKED"         ~ "SAMOA",
    team_norm == "TONGAUNRANKED"         ~ "TONGA",
    team_norm == "ERITREAUNRANKED"       ~ "ERITREA",
    team_norm == "ERITREA"               ~ "ERITREA",   # déjà bon mais absent de results
    
    # ── Typo ─────────────────────────────────────────────────────────────────
    team_norm == "CAMPOON"              ~ "CAMEROON",
    team_norm == "CURACAO"              ~ "CURACAO",   # déjà bon
    team_norm == "CURAAAO"              ~ "CURACAO",

    
    TRUE ~ team_norm
  ))
```

## Formatting the ranking 

```{r}
ranking_clean <- ranking %>% 
  mutate(rank_points = total_points,
         periods = 31,
         team = team_norm) %>% 
  filter(!is.na(rank_points)) %>% 
  select(periods, team, rank_points)
```

```{r}
ranking_worldcup <- ranking_clean %>% 
  filter(team %in% unique(c(final_data$home_team, final_data$away_team)))
```

```{r}
ranking_ready <- data.frame(
  team        = as.character(ranking_periods$team),
  rank_points = as.numeric(ranking_periods$rank_points),
  periods     = as.integer(ranking_periods$periods),
  stringsAsFactors = FALSE
)
```

# Coordination between the initialization and data teams

```{r}
teams_new_model <- unique(c(final_data$home_team, final_data$away_team))  

cat("Nb équipes dans le nouveau modèle :", length(teams_new_model), "\n")
cat("Nb équipes dans teams_kept (posteriors filtrés) :", length(teams_kept), "\n")
```

```{r}
only_in_new   <- setdiff(teams_new_model, teams_kept)
only_in_kept  <- setdiff(teams_kept, teams_new_model)

if (length(only_in_new) > 0) {
  cat("Équipes dans le nouveau modèle SANS prior disponible :\n")
  print(only_in_new)
}
if (length(only_in_kept) > 0) {
  cat("Équipes avec un prior extrait mais ABSENTES du nouveau modèle :\n")
  print(only_in_kept)
}

if (length(only_in_new) == 0 && length(only_in_kept) == 0) {
  cat("Correspondance parfaite entre les deux ensembles d'équipes.\n")
} else {
  stop("Les ensembles d'équipes ne correspondent pas exactement -- corrigez avant de continuer.")
}
```

```{r}
reorder_idx <- match(teams_new_model, teams_kept)

inits_pred2 <- lapply(inits_pred, function(init_c) {
  init_c$att_raw_centered <- init_c$att_raw_centered[reorder_idx]
  init_c$def_raw_centered <- init_c$def_raw_centered[reorder_idx]
  init_c
})
```

# The predictions 

```{r}
diag_infl_predictions <- stan_foot(
  data = final_data,
  model = "diag_infl_biv_pois",
  predict = 16,
  ranking = ranking_worldcup,
  init = inits_pred2,
  home_effect = FALSE,
  chains = 4, parallel_chains = 4,
  iter_sampling = 1000,
  iter_warmup = 4000,  
  adapt_delta = 0.995, 
  max_treedepth = 14, 
  seed = 123
)
```

```{r}
prob_diag_infl_pred <- foot_prob(object = diag_infl_predictions, data = final_data)
```

```{r}
colnames(prob_diag_infl_pred$prob_table) <- c(
      "home",
      "away",
      "home win",
      "draw",
      "away win",
      "mlo"
    )

prob_diag_infl_pred$prob_table
```

```{r}
write.csv(prob_diag_infl_pred$prob_table, 
          file = "~/work/World-Cup-2026/results/round_of_32_diag_infl_biv_pois.csv", 
          row.names = FALSE)
```

```{r}
new_names <- c(
  "MEXICO"          = "Mexico",
  "SOUTHKOREA"      = "South Korea",
  "CANADA"            = "Canada",
  "UNITEDSTATES"      = "United States",
  "QATAR"      = "Qatar",
  "BRAZIL"           = "Brazil",
  "HAITI"            = "Haiti",
  "AUSTRALIA"    = "Australia",
  "GERMANY"      = "Germany",
  "IVORYCOAST"          = "Ivory Coast",
  "NETHERLANDS"            = "Netherlands",
  "SWEDEN"          = "Sweden",
  "BELGIUM"           = "Belgium",
  "IRAN"          = "Iran",
  "SPAIN"           = "Spain",
  "SAUDIARABIA"           = "Saudi Arabia",
  "FRANCE"          = "France",
  "IRAQ"          = "Iraq", 
  "ARGENTINA"          = "Argentina",
  "AUSTRIA"        = "Austria",
  "PORTUGAL"           = "Portugal",
  "UZBEKISTAN"      = "Uzbekistan",
  "ENGLAND"   = "England",
  "GHANA"         = "Ghana",
  "CZECHREPUBLIC"         = "Czech Republic",
  "SWITZERLAND"          = "Switzerland",
  "SCOTLAND"           = "Scotland",
  "TURKEY"         = "Turkey",
  "ECUADOR"            = "Ecuador",
  "TUNISIA"        = "Tunisia",
  "NEWZEALAND"          = "New Zealand",
  "URUGUAY"         = "Uruguay",
  "NORWAY"       = "Norway",
  "JORDAN"          = "Jordan",
  "COLOMBIA"     = "Colombia",
  "PANAMA"          = "Panama",
  "SOUTHAFRICA"     = "South Africa",
  "BOSNIAANDHERZEGOVINA"           = "Bosnia and Herzegovina",
  "MOROCCO"             = "Morocco",
  "PARAGUAY"       = "Paraguay",
  "CURACAO" = "Curacao",
  "JAPAN" = "Japan",
  "EGYPT" = "Egypt",
  "CAPEVERDE" = "Cape Verde",
  "SENEGAL" = "Senegal",
  "ALGERIA" = "Algeria",
  "DRCONGO" = "DR Congo",
  "CROATIA" = "Croatia"
)

p_32 <- prob_diag_infl_pred$prob_plot

for (old in names(new_names)) {
  p_32$data$new_matches <- gsub(old, new_names[old], p_32$data$new_matches, ignore.case = TRUE)
}

p_32$theme$axis.text.x$size <- 10
p_32$theme$axis.text.y$size <- 10
p_32$theme$axis.title$size  <- 10
p_32$theme$strip.text$size  <- 10 

p_32 <- p_32 + theme(
  strip.text = element_text(size = 10, face = "bold"),  
  axis.text  = element_text(size = 10),                 
  axis.title = element_text(size = 10, face = "bold"), 
  plot.title = element_text(size = 14, face = "bold") 
)

p_32 <- p_32 + guides(color = "none")

p_32
```

```{r}
ggsave("~/work/World-Cup-2026/results/plot_diag_infl_round_of_32.png", plot = p_32, width = 20, height = 15, dpi = 300)
```



