---
title: "Predictions groupstage2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The theory

We use a **diagonal-inflated Bivariate-Poisson model with dynamic team-specific abilities** for the attack and the defence. Let $(X_{i}, Y_{i})$ denote the random number of goals scored by the home and the away team in the $i$-th game, $i=1,\ldots,n$, respectively. $\mathsf{ranking}$ denotes the latest Coca-Cola FIFA ranking available for each team prior to the match being played, whereas att and def denote the attack and the defence abilities, respectively.

\begin{align}
(X_i, Y_i) &\ \sim \ \begin{cases} (1-p) \text{BP}(x_i, y_i |\lambda_1, \lambda_2, \lambda_3) \ \ \ & \text{if} \    x \ne y \\ (1-p) \text{BP}(x_i, y_i | \lambda_1, \lambda_2, \lambda_3) + pD(x, \eta) \ \ \ & \text{if} \   x = y, \end{cases}, \\
\log(\lambda_{1i}) &=\    \text{att}_{h_i, t}+ \text{def}_{a_i,t} + \frac{\gamma}{2}(\mathsf{ranking}_{h_i,t}-\mathsf{ranking}_{a_i,t}) \\
\log(\lambda_{2i}) & =\    \text{att}_{a_i,t} + \text{def}_{h_i,t} - \frac{\gamma}{2}(\mathsf{ranking}_{h_i,t}-\mathsf{ranking}_{a_i}), \ \ i=1,\ldots,n\ (\text{matches}), \\
\log(\lambda_{3i}) & =\ \rho,\\
\text{att}_{k, t}&  \sim \ \mathcal{N}(\text{att}_{k, t-1}, \sigma^2), \\
\text{def}_{k, t} & \sim \  \mathcal{N}(\text{def}_{k, t-1}, \sigma^2),\\
\rho, \ \gamma & \sim \mathcal{N}(0,1) \\
p & \sim \text{Uniform}(0,1)\\
& \sum_{k=1}^{n_t} \text{att}_{k, }=0, \  \sum_{k=1}^{n_t}\text{def}_{k, }=0, \ \ k=1,\ldots n_t \ (\text{teams}), \  t=1,\ldots, T \ (\text{times}).
\label{eq:scoring_rue}
\end{align}

This model is designed to predict football match scores, where $(X_i, Y_i)$ represent the number of goals scored by the home team and the away team, respectively, in match i. 

Model fitting has been obtained through the Hamiltonian Monte Carlo sampling, 2000 iterations, 4 chains using the ```footBayes``` \texttt{R} package (with the underlying ```rstan``` package). 

The historical data used to fit the models come from *all the international matches played since the 1st of January 2022*, both for the games and the ranking (the oldest ranking used is the one of the 23rd of December 2021).

The idea is to provide a dynamic predictive scenario: at the end of each match-day, the model will be refitted  to predict the remaining matches. So for now, we have started by making predictions only for the first matchday of the group stage. 

# The data

## The results of groupstage1

```{r}
results_groupstage1 <- data.frame(
  periods    = rep(31, 24),   
  home_team  = c("MEXICO","SOUTHKOREA","CANADA","UNITEDSTATES",
                 "QATAR","BRAZIL","HAITI","AUSTRALIA",
                 "GERMANY", "NETHERLANDS","IVORYCOAST","SWEDEN",
                 "SPAIN","BELGIUM","SAUDIARABIA","IRAN",
                 "FRANCE","IRAQ","ARGENTINA","AUSTRIA",
                 "PORTUGAL","ENGLAND","GHANA","UZBEKISTAN"),
  home_goals = c(2,2,1,4,1,1,0,2,
                 7,2,1,5,NA,NA,NA,NA,
                 NA,NA,NA,NA,NA,NA,NA,NA), 
  away_team  = c("SOUTHAFRICA","CZECHREPUBLIC","BOSNIAANDHERZEGOVINA","PARAGUAY",
                 "SWITZERLAND","MOROCCO","SCOTLAND","TURKEY",
                 "CURACAO", "JAPAN","ECUADOR","TUNISIA",
                 "CAPEVERDE","EGYPT","URUGUAY","NEWZEALAND",
                 "SENEGAL","NORWAY","ALGERIA","JORDAN",
                 "DRCONGO","CROATIA","PANAMA","COLOMBIA"),
  away_goals = c(0,1,1,1,1,1,1,0,
                 1,2,0,1,NA,NA,NA,NA,
                 NA,NA,NA,NA,NA,NA,NA,NA)
)
```

## Games of groupstage2

```{r}
games_groupstage2 <- data.frame(
  periods    = rep(31,24),
  home_team  = c("CZECHREPUBLIC","SWITZERLAND","CANADA","MEXICO",
                 "UNITEDSTATES","SCOTLAND","BRAZIL","TURKEY",
                 "NETHERLANDS","GERMANY","ECUADOR","TUNISIA",
                 "SPAIN","BELGIUM","URUGUAY","NEWZEALAND",
                 "ARGENTINA","FRANCE","NORWAY","JORDAN",
                 "PORTUGAL","ENGLAND","PANAMA","COLOMBIA"),
  home_goals = rep(NA,24),
  away_team  = c("SOUTHAFRICA","BOSNIAANDHERZEGOVINA","QATAR","SOUTHKOREA",
                 "AUSTRALIA","MOROCCO","HAITI","PARAGUAY",
                 "SWEDEN","IVORYCOAST","CURACAO","JAPAN",
                 "SAUDIARABIA","IRAN","CAPEVERDE","EGYPT",
                 "AUSTRIA","IRAQ","SENEGAL","ALGERIA",
                 "UZBEKISTAN","GHANA","CROATIA","DRCONGO"),
  away_goals = rep(NA,24)
)
```

## Final groupstage2

```{r}
groupstage2 <- rbind(
  groupstage1[1:(nrow(groupstage1) - 24), ],  
  results_groupstage1,
  games_groupstage2
)
```


# The predictions

## Initialisation of the paramters

```{r}
# Extraction of the postreriors
draws_df <- diag_infl_j1$fit$draws(format = "draws_df")

n_chains <- 4

inits_j2 <- lapply(1:n_chains, function(c) {
  draw_idx <- sample(nrow(draws_df), 1)
  
  list(
    rho    = as.numeric(draws_df[draw_idx, "rho"]),
    gamma  = as.numeric(draws_df[draw_idx, "gamma"]),
    p      = as.numeric(draws_df[draw_idx, "p"]),
    att    = as.numeric(draws_df[draw_idx, grep("^att\\[", colnames(draws_df))]),
    def    = as.numeric(draws_df[draw_idx, grep("^def\\[", colnames(draws_df))])
  )
})
```

## The predictions

```{r}
groupstage2_teams <- unique(c(groupstage2$home_team, groupstage2$away_team))

ranking_footBayes_clean2 <- ranking_footBayes %>%
  filter(team %in% groupstage2_teams)
```

```{r}
diag_infl_j2 <- stan_foot(
  data          = groupstage2,
  model         = "diag_infl_biv_pois",
  predict       = 24,              
  ranking       = ranking_footBayes_clean2,
  dynamic_type  = "seasonal",
  init          = inits_j2,       
  home_effect   = FALSE,
  chains        = 4,
  iter_sampling = 1000,
  iter_warmup   = 500,            
  seed          = 456
)
```

```{r}
prob_diag_infl_j2 <- foot_prob(object = diag_infl_j2, data = groupstage2)
```

```{r}
colnames(prob_diag_infl_j2$prob_table) <- c(
      "home",
      "away",
      "home win",
      "draw",
      "away win",
      "mlo"
    )
```

```{r}
write.csv(prob_diag_infl_j2$prob_table, 
          file = "results/groupstage2_diag_infl_biv_pois.csv", 
          row.names = FALSE)
```

```{r}
new_names <- c(
  "MEXICO"          = "Mexico",
  "SOUTHKOREA"      = "South Korea",
  "CANADA"            = "Canada",
  "UNITEDSTATES"      = "United States",
  "QATAR"      = "Qatar",
  "BRAZIL"           = "Brazil",
  "HAITI"            = "Haiti",
  "AUSTRALIA"    = "Australia",
  "GERMANY"      = "Germany",
  "IVORYCOAST"          = "Ivory Coast",
  "NETHERLANDS"            = "Netherlands",
  "SWEDEN"          = "Sweden",
  "BELGIUM"           = "Belgium",
  "IRAN"          = "Iran",
  "SPAIN"           = "Spain",
  "SAUDIARABIA"           = "Saudi Arabia",
  "FRANCE"          = "France",
  "IRAQ"          = "Iraq", 
  "ARGENTINA"          = "Argentina",
  "AUSTRIA"        = "Austria",
  "PORTUGAL"           = "Portugal",
  "UZBEKISTAN"      = "Uzbekistan",
  "ENGLAND"   = "England",
  "GHANA"         = "Ghana",
  "CZECHREPUBLIC"         = "Czech Republic",
  "SWITZERLAND"          = "Switzerland",
  "SCOTLAND"           = "Scotland",
  "TURKEY"         = "Turkey",
  "ECUADOR"            = "Ecuador",
  "TUNISIA"        = "Tunisia",
  "NEWZEALAND"          = "New Zealand",
  "URUGUAY"         = "Uruguay",
  "NORWAY"       = "Norway",
  "JORDAN"          = "Jordan",
  "COLOMBIA"     = "Colombia",
  "PANAMA"          = "Panama",
  "SOUTHAFRICA"     = "South Africa",
  "BOSNIAANDHERZEGOVINA"           = "Bosnia and Herzegovina",
  "MOROCCO"             = "Morocco",
  "PARAGUAY"       = "Paraguay",
  "CURACAO" = "Curacao",
  "JAPAN" = "Japan",
  "EGYPT" = "Egypt",
  "CAPEVERDE" = "Cape Verde",
  "SENEGAL" = "Senegal",
  "ALGERIA" = "Algeria",
  "DRCONGO" = "DR Congo",
  "CROATIA" = "Croatia"
)

p_j2 <- prob_diag_infl_j2$prob_plot

for (old in names(new_names)) {
  p_j2$data$new_matches <- gsub(old, new_names[old], p_j2$data$new_matches, ignore.case = TRUE)
}

p_j2$theme$axis.text.x$size <- 10
p_j2$theme$axis.text.y$size <- 10
p_j2$theme$axis.title$size  <- 10
p_j2$theme$strip.text$size  <- 10 

p_j2 <- p_j2 + theme(
  strip.text = element_text(size = 10, face = "bold"),  
  axis.text  = element_text(size = 10),                 
  axis.title = element_text(size = 10, face = "bold"), 
  plot.title = element_text(size = 14, face = "bold") 
)

p_j2 <- p_j2 + guides(color = "none")
```

```{r}
ggsave("results/plot_diag_infl_j2.png", plot = p_j2, width = 20, height = 15, dpi = 300)
```


