---
title: "Last year"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Packages needed
library(dplyr)
library(tidyr)
library(loo)
library(cmdstanr)
library(footBayes)
library(posterior)
library(bayesplot)
library(instantiate)
library(stringi)
library(stringr)
library(readr)
```

# The theory

We use a **diagonal-inflated Bivariate-Poisson model with dynamic team-specific abilities** for the attack and the defence. Let $(X_{i}, Y_{i})$ denote the random number of goals scored by the home and the away team in the $i$-th game, $i=1,\ldots,n$, respectively. $\mathsf{ranking}$ denotes the latest Coca-Cola FIFA ranking available for each team prior to the match being played, whereas att and def denote the attack and the defence abilities, respectively.

\begin{align}
(X_i, Y_i) &\ \sim \ \begin{cases} (1-p) \text{BP}(x_i, y_i |\lambda_1, \lambda_2, \lambda_3) \ \ \ & \text{if} \    x \ne y \\ (1-p) \text{BP}(x_i, y_i | \lambda_1, \lambda_2, \lambda_3) + pD(x, \eta) \ \ \ & \text{if} \   x = y, \end{cases}, \\
\log(\lambda_{1i}) &=\    \text{att}_{h_i, t}+ \text{def}_{a_i,t} + \frac{\gamma}{2}(\mathsf{ranking}_{h_i,t}-\mathsf{ranking}_{a_i,t}) \\
\log(\lambda_{2i}) & =\    \text{att}_{a_i,t} + \text{def}_{h_i,t} - \frac{\gamma}{2}(\mathsf{ranking}_{h_i,t}-\mathsf{ranking}_{a_i}), \ \ i=1,\ldots,n\ (\text{matches}), \\
\log(\lambda_{3i}) & =\ \rho,\\
\text{att}_{k, t}&  \sim \ \mathcal{N}(\text{att}_{k, t-1}, \sigma^2), \\
\text{def}_{k, t} & \sim \  \mathcal{N}(\text{def}_{k, t-1}, \sigma^2),\\
\rho, \ \gamma & \sim \mathcal{N}(0,1) \\
p & \sim \text{Uniform}(0,1)\\
& \sum_{k=1}^{n_t} \text{att}_{k, }=0, \  \sum_{k=1}^{n_t}\text{def}_{k, }=0, \ \ k=1,\ldots n_t \ (\text{teams}), \  t=1,\ldots, T \ (\text{times}).
\label{eq:scoring_rue}
\end{align}

This model is designed to predict football match scores, where $(X_i, Y_i)$ represent the number of goals scored by the home team and the away team, respectively, in match i. 

Model fitting has been obtained through the Hamiltonian Monte Carlo sampling, 2000 iterations, 4 chains using the ```footBayes``` \texttt{R} package (with the underlying ```rstan``` package). 

The historical data used to fit the models come from *all the international matches played since the 1st of January 2022*, both for the games and the ranking (the oldest ranking used is the one of the 23rd of December 2021).

The idea is to provide a dynamic predictive scenario: at the end of each match-day, the model will be refitted  to predict the remaining matches. So for now, we have started by making predictions only for the first matchday of the group stage. 

# The data

## Data loading and preprocessing

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("~/work/World-Cup-2026/codes/ranking_preprocessing_last_year.R")
```

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("~/work/World-Cup-2026/codes/ranking2025_11_19_preprocessing.R")
```

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("~/work/World-Cup-2026/codes/ranking2026_06_11_preprocessing.R")
```

```{r, include=FALSE}
# Exécute le script et charge tout son contenu
source("~/work/World-Cup-2026/codes/results_preprocessing_last_year.R")
```

```{r}
ranking <- rbind(ranking, ranking2025_11_19, ranking2026_06_11)
```


## All the teams are in the ranking ?

```{r}
all_rank_norms <- unique(ranking$team_norm)
res_teams <- unique(c(results$home_norm, results$away_norm))

unmatched_in_results <- setdiff(res_teams, all_rank_norms)

cat("\n=== Teams dans results sans correspondance dans ranking ===\n")
print(sort(unmatched_in_results))
```

## All the teams in ranking are in results 

```{r}
all_rank_norms <- unique(ranking$team_norm)
res_teams <- unique(c(results$home_norm, results$away_norm))

unmatched_in_results <- setdiff(res_teams, all_rank_norms)

cat("\n=== Teams dans results sans correspondance dans ranking ===\n")
print(sort(unmatched_in_results))
```

## Correction of the names which cause problems 

```{r}
ranking <- ranking |>
  mutate(team_norm = case_when(
    
    # ── Renommages officiels ──────────────────────────────────────────────────
    team_norm == "CZECHIA"              ~ "CZECHREPUBLIC",
    team_norm == "AOTEAROANEWZEALAND"   ~ "NEWZEALAND",
    team_norm == "THEGAMBIA"            ~ "GAMBIA",
    team_norm == "TURKIYE"              ~ "TURKEY",
    team_norm == "TA14RKIYE"            ~ "TURKEY",   # encodage corrompu
    
    # ── Variantes de noms ────────────────────────────────────────────────────
    team_norm == "CABOVERDE"            ~ "CAPEVERDE",
    team_norm == "CAPEVERDEISLANDS"     ~ "CAPEVERDE",
    team_norm == "COTEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CATEDIVOIRE"          ~ "IVORYCOAST",
    team_norm == "CONGODR"              ~ "DRCONGO",
    team_norm == "KOREAREPUBLIC"        ~ "SOUTHKOREA",
    team_norm == "KOREADPR"             ~ "NORTHKOREA",
    team_norm == "KYRGYZREPUBLIC"       ~ "KYRGYZSTAN",
    team_norm == "USA"                  ~ "UNITEDSTATES",
    team_norm == "USVIRGINISLANDS"      ~ "UNITEDSTATESVIRGINISLANDS",
    team_norm == "STKITTSANDNEVIS"      ~ "SAINTKITTSANDNEVIS",
    team_norm == "STLUCIA"              ~ "SAINTLUCIA",
    team_norm == "STVINCENTANDTHEGRENADINES" ~ "SAINTVINCENTANDTHEGRENADINES",
    team_norm == "BRUNEIDARUSSALAM"     ~ "BRUNEI",
    team_norm == "CHINESETAIPEI"        ~ "TAIWAN",
    team_norm == "IRIRAN"               ~ "IRAN",   # encodage corrompu
    
    # ── Variantes São Tomé ───────────────────────────────────────────────────
    team_norm == "SAOTOMEANDPRINCIPE"   ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMEEPRINCIPE"     ~ "SAOTOMEETPRINCIPE",
    team_norm == "SAOTOMACANDPRANCIPE"  ~ "SAOTOMEETPRINCIPE",
    
    # ── Suffixes (unranked) → nom de base ────────────────────────────────────
    team_norm == "AMERICANSAMOAUNRANKED" ~ "AMERICANSAMOA",
    team_norm == "SAMOAUNRANKED"         ~ "SAMOA",
    team_norm == "TONGAUNRANKED"         ~ "TONGA",
    team_norm == "ERITREAUNRANKED"       ~ "ERITREA",
    team_norm == "ERITREA"               ~ "ERITREA",   # déjà bon mais absent de results
    
    # ── Typo ─────────────────────────────────────────────────────────────────
    team_norm == "CAMPOON"              ~ "CAMEROON",
    team_norm == "CURACAO"              ~ "CURACAO",   # déjà bon
    team_norm == "CURAAAO"              ~ "CURACAO",

    
    TRUE ~ team_norm
  ))
```

```{r}
results <- results |>
  mutate(
    home_norm = case_when(
      home_norm == "CURAAAO"             ~ "CURACAO",
      home_norm == "SAOTOMACANDPRANCIPE" ~ "SAOTOMEETPRINCIPE",
      TRUE ~ home_norm
    ),
    away_norm = case_when(
      away_norm == "CURAAAO"             ~ "CURACAO",
      away_norm == "SAOTOMACANDPRANCIPE" ~ "SAOTOMEETPRINCIPE",
      TRUE ~ away_norm
    )
  )
```

## Formatting the ranking

```{r}
ranking_clean <- ranking %>% 
  mutate(rank_points = total_points,
         date = as.Date(date)) %>% 
  filter(!is.na(rank_points)) %>% 
  select(date, team_norm, rank_points) %>% 
  mutate(team = team_norm) %>% 
  select(date, team, rank_points)
```

```{r}
top_140 <- ranking_clean %>% 
  filter(date == max(date, na.rm = TRUE)) %>% 
  slice_max(order_by = rank_points, n = 180, with_ties = FALSE) %>% 
  pull(team)
```

```{r}
ranking_worldcup <- ranking_clean %>% 
  filter(team %in% top_140)
```

```{r}
ranking_worldcup <- ranking_worldcup %>% 
  arrange(date, rank_points)
```

## Creation of the periods

```{r}
ranking_periods <- ranking_worldcup %>%
  arrange(date) %>%
  mutate(periods = dense_rank(date))

dates_periods <- ranking_periods %>%
  select(date, periods) %>%
  distinct()
```

```{r}
worldcup2026_periods <- results %>%
  full_join(dates_periods, by = c("date" = "date")) %>%
  arrange(date) %>%
  fill(periods, .direction = "down") %>% 
  filter(!is.na(home_team))
```

## Ranking ready to be used by footBayes package

```{r}
ranking_footBayes <- data.frame(
  team        = as.character(ranking_periods$team),
  rank_points = as.numeric(ranking_periods$rank_points),
  periods     = as.integer(ranking_periods$periods),
  stringsAsFactors = FALSE
)
```

```{r}
n_periodes <- n_distinct(ranking_footBayes$periods)
groupstage2_teams <- unique(c(results$home_norm, results$away_norm))

equipes_completes <- ranking_footBayes %>%
  group_by(team) %>%
  summarise(n_periodes_team = n_distinct(periods)) %>%
  filter(n_periodes_team == n_periodes) %>%
  pull(team)

ranking_footBayes_clean2 <- ranking_footBayes %>%
  filter(team %in% groupstage2_teams,
         team %in% equipes_completes)
```

## Formatting of results

```{r}
worldcup2026 <- worldcup2026_periods %>%
  mutate(date = as.Date(date)) %>%
  mutate(is_future = is.na(home_score) | is.na(away_score)) %>%
  arrange(is_future, date)

summary(worldcup2026)
table(worldcup2026$is_future)
```

```{r}
worldcup2026 <- worldcup2026 %>%
  filter((home_norm %in% ranking_footBayes$team) & (away_norm %in% ranking_footBayes$team) )
```

```{r}
friendly_teams <- worldcup2026 %>%
  filter(tournament == "Friendly") %>%
  select(home_norm, away_norm) %>%
  pivot_longer(
    cols = c(home_norm, away_norm),
    names_to = "side",
    values_to = "team"
  ) %>%
  distinct(team) %>%
  arrange(team)

wc48_teams <- worldcup2026 %>%
  filter(is_future == TRUE) %>%
  select(home_norm, away_norm)

wc48_teams <- unique(c(wc48_teams$home_norm, wc48_teams$away_norm))

length(wc48_teams)
```

```{r}
# The training data (the games played since the 1st of January of 2022)
played_matches <- worldcup2026 %>%
  filter(is_future == FALSE)

# Which games have missing values ?
played_matches_with_na <- played_matches %>%
  filter(if_any(everything(), is.na))

# How much games with missing values do we have ?
nrow(played_matches_with_na)
```


```{r}
# All the games with "suspicious" outcomes
worldcup2026_flagged <- worldcup2026 %>%
  mutate(
    total_goals = ifelse(is_future == FALSE, home_score + away_score, NA_real_),
    goal_diff   = ifelse(is_future == FALSE, abs(home_score - away_score), NA_real_),

    both_teams_in_wc48 = (home_norm %in% wc48_teams) &
      (away_norm %in% wc48_teams),

    friendly_suspect = (tournament == "Friendly") &
      (is_future == FALSE) &
      (
        goal_diff >= 5 |
          (total_goals >= 7 & goal_diff >= 4)
      ),

    remove_match = friendly_suspect & !both_teams_in_wc48
  )

# Friendly games with suspicious resultas 
friendly_suspects <- worldcup2026_flagged %>%
  filter(friendly_suspect == TRUE) %>%
  select(
    periods,
    date,
    home_team,
    home_norm,
    away_team,
    away_norm,
    home_score,
    away_score,
    tournament,
    total_goals,
    goal_diff,
    both_teams_in_wc48,
    remove_match,
    neutral
  ) %>%
  arrange(desc(goal_diff), desc(total_goals))

# Then we keep only the non suspicious games 
worldcup2026_clean <- worldcup2026_flagged %>%
  filter(remove_match == FALSE)
```

```{r}
cat("Matches before cleaning:", nrow(worldcup2026), "\n")
cat("Matches after cleaning:", nrow(worldcup2026_clean), "\n")
cat("Removed matches:", nrow(worldcup2026) - nrow(worldcup2026_clean), "\n")

table(worldcup2026$is_future)
table(worldcup2026_clean$is_future)

summary(worldcup2026_clean)
```

## Results ready for footBayes package

```{r}
results_footBayes <- worldcup2026_clean %>%
  arrange(is_future, date) %>%
  rename(
    home_goals = home_score,
    away_goals = away_score
  ) %>%
  select(
    periods,
    home_norm,
    away_norm,
    home_goals,
    away_goals,
    is_future
  ) %>% 
  mutate(home_team = home_norm,
         away_team = away_norm)
```

```{r}
n_future <- sum(results_footBayes$is_future)

cat("Number of future matches:", n_future, "\n")

if (n_future != 72) {
  warning("The number of future matches is not 72 after filtering.")
}

n_total <- nrow(results_footBayes)
n_train <- n_total - n_future

results_footBayes_ready <- results_footBayes %>% 
  select(periods, home_team, home_goals, away_team, away_goals)
```

```{r}
equipes_ranking <- unique(ranking_footBayes_clean2$team)

results_footBayes_ready <- results_footBayes_ready %>%
  filter(home_team %in% equipes_ranking,
         away_team %in% equipes_ranking)
```

## Which are not in both home and away ?

```{r}
# Teams only in home and not in away
teams_home_only <- setdiff(results_footBayes_ready$home_team, results_footBayes_ready$away_team)

# Teams only in away and not in home
teams_away_only <- setdiff(results_footBayes_ready$away_team, results_footBayes_ready$home_team)

# Games concerned
matches_home_only <- results_footBayes_ready[results_footBayes_ready$home_team %in% teams_home_only, ]
matches_away_only <- results_footBayes_ready[results_footBayes_ready$away_team %in% teams_away_only, ]

matches_home_only
matches_away_only
```

## How to fix the problem ?

```{r}
to_remove <- data.frame(
  home_team = c("MALAYSIA","UNITEDSTATES","VIETNAM","PAPUANEWGUINEA","TURKMENISTAN","ANTIGUAANDBARBUDA","NEWCALEDONIA","PUERTORICO","GUYANA","THAILAND","INDONESIA","CUBA","SAINTKITTSANDNEVIS","COMOROS","MALDIVES","DOMINICANREPUBLIC","AZERBAIJAN"),
  away_team = c("SINGAPORE","SOUTHKOREA","NEPAL","SOLOMONISLANDS","TAIWAN","GUYANA","JAMAICA","SAUDIARABIA","BELIZE","FIJI","LEBANON","SAINTLUCIA","BELIZE","YEMEN","AFGHANISTAN","CUBA","SAINTLUCIA"),
  periods  = c(1,1,2,3,3,3,6,7,6,1,1,3,3,4,7,6,6)
)
```

```{r}
should_remove <- function(df, to_remove) {
  flag <- rep(FALSE, nrow(df))
  for (i in seq_len(nrow(to_remove))) {
    match_idx <- df$home_team == to_remove$home_team[i] &
                 df$away_team == to_remove$away_team[i] &
                 df$periods >= to_remove$periods[i] 
    flag <- flag | match_idx
  }
  flag
}

rows_flagged <- should_remove(results_footBayes_ready, to_remove)
```

```{r}
results_footBayes_ready <- results_footBayes_ready[!rows_flagged, ]
```

```{r}
inversed_games <- data.frame(
  periods  = c(1,1,2,3,1,1,6,2),
  home_team = c("SINGAPORE","SOUTHKOREA","NEPAL","SOLOMONISLANDS","FIJI","LEBANON","CUBA","VIETNAM"),
  home_goals = c(1,2,0,0,0,0,1,3),
  away_team = c("MALAYSIA","UNITEDSTATES","VIETNAM","PAPUANEWGUINEA","THAILAND","INDONESIA","DOMINICANREPUBLIC","NEPAL"),
  away_goals = c(2,0,1,1,3,0,1,1)
)


results_footBayes_ready <- rbind(results_footBayes_ready, inversed_games)
```

```{r}
results_footBayes_ready <- results_footBayes_ready %>% arrange(periods)
```

